---
title: "International Economics dataset, association rules"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
author: "Mateusz Baryła"
date: "6 12 2020"
bibliography: bibliography.bibtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      results = 'asis')
pacman::p_load(tidyverse, kableExtra, kohonen, arules)

knitr::opts_chunk$set(plain.ascii = FALSE,        # Always use this option in Rmd documents
           style        = "rmarkdown", # Always use this option in Rmd documents
           footnote     = NA,          # Makes html-rendered results more concise
           subtitle.emphasis = FALSE)  # Improves layout with some rmardown themes
```

## Introduction

Association rules are often used for reducing number of transactions in the 
databases where each row is coded by binary attributes. There is no need for 
them to be trained, or labeling dataset beforehand. Most common applications 
are found in the market basket analysis, discovering interesting patterns of 
DNA and protein sequences, common patterns of behavior can be found for 
customers that proceed customers dropping their cell phone operator.

## Dataset

### Instacart Market Basket Analysis

```{r}
aisles <- read.csv("data\\aisles.csv", sep=",")
departments <- read.csv("data\\departments.csv", sep=",")
order_products_train <- read.csv("data\\order_products__prior.csv", sep=",")
orders <- read.csv("data\\orders.csv", sep=",")
products <- read.csv("data\\products.csv", sep=",")
```

```{r}
trans <- 
  order_products_train %>% 
  inner_join(products, by = "product_id", keep = FALSE) %>%
  inner_join(aisles, by = "aisle_id", keep = FALSE) %>% 
  inner_join(orders, by = "order_id", keep = FALSE) %>% 
  inner_join(departments, by ="department_id", keep = FALSE)
```

```{r}
trans %>% 
  select(-order_id,-aisle_id, -department_id, -user_id, -eval_set) %>% 
  head() %>%
  write.csv("E:\\GitHub\\unsupervised-learning\\data\\transactions.csv", 
                    sep=",", row.names = FALSE, quote = TRUE)
```

Data provided on kaggle consist of 6 different files, which will be analyzed and 
merged into one big transactional dataset after basic preprocessing and 
EDA process. Products prior are merged into one dataset with 14 columns. The 
single observation consist of order_id, product_name, add_to_cart_order, reordered,
eval_set, order_dow, order_hour_of_day, days_since_prior_order, department_name.
Norows from original dataframe are removed therefore 32434489 observations will be
investigated.

```{r}
trans <- read.transactions(file = "E:\\GitHub\\unsupervised-learning\\data\\transactions.csv",format = "basket", sep = ",", quote = "")
```

### The nature of the dataset

In order to provide master dataset that can be used by researchers from
international political economy 89 data resources have been merged. Observations
are identified by country-year the unit of analysis. Countries are identified by 
Gleditsch-Ward number ot in an alternative version by Correlates of War (COW).
Most of the dataset components begin after Second World War. 

### Variable naming

Each of 89 dataset has been given a unique suffix that uniquely identifies it.

## Dataset preprocessing

Firstly data is loaded, then we convert it to tibble as tidyverse will be 
mainly used in this project.

```{r}
load(paste0("data/", "master_ipe_v4.rdata"))
ipe_v4 <- as_tibble(ipe_v4)
```

Dimensions of the dataset needs to be investigated as well.

```{r}
ipe_v4 %>% dim()
```


Master table consists of 25850 rows and 1043 variables. Each of them is described 
in the codebook provided by library maintainers. They are divided into following 
groups:
* economic
* political
* social and cultural
* geographic 
* other (e.g. infrastructure, military)

Very good quality of the data is provided by Varities of Democracy Dataset (VDEM)
inside political section, it will definitely be included not only because of 
the quality but also its popularity and inside class recommendation. 

The other  dataset chosen by me is going to be "Polity IV Democracy".

```{r}
pol_ds <- 
  ipe_v4 %>% 
  filter(country == "Poland", year > 1918, 
         !(year %in% c(1939, 1940, 1941, 1942, 1943, 1944, 1945, 2020))) %>% 
  select(year, starts_with("v2"), contains("P4"), 
         -c(change_P4, sf_P4, fragment_P4, regtrans_P4, countryname_raw_P4, 
            v2x_hosinter_VDEM, v2x_suffr_VDEM, v2elsrgel_VDEM, v2elreggov_VDEM,
            v2xlg_leginter_VDEM, v2x_elecreg_VDEM, v2xlg_elecreg_VDEM)) 
pol_ds %>%
  head() %>% 
  kable()
```

## Exploratory data analysis

In every modeling process we should perform exploratory data analysis before moving to 
the next points. 

```{r}
dfSummary(pol_ds, plain.ascii = FALSE, style = "grid", 
          graph.magnif = 0.75, valid.col = FALSE, tmp.img.dir = "/tmp")
```

Thanks to the broad landscape of autoEDA packages we can very easily generate 
some intuition towards the analyzed data. The chosen subset of data is associated 
with brilliant starting point for analysis having only two columns,
v2x_gender_VDEM and v2x_genpp_VDEM, with missing value. The graphs show distribution
for each variable. The year column shows that there is no data between years 1939-1948.
Most of the variables there are so called indexes with range between 0 and 1.
For P4 varialbes the minimum value can be lower than zero. Durable_P4 variable is 
highly skewed. For v2xcl_rol_VDEM we can clearly see oposite pattern, however  
there is a misssing range of numbers for which we have little to no observations. 

# Association rules

## General view



## Measuring rule interest

Two metrics are taken into account support 
and confidence for association rules. Support describes how frequently an items or rule occurs in the data. THis is simply the number of occurrences of particular item
divided by total number of rows.

Confidence measure gives the intuition what is the proportion of transactions where the presence of item or itemset X results in the presence of item or itemset Y.

Combination of two mentioned metrics is lift which is described as the division of confidence by support. A value around 1 suggests independent itemsets. Given a value less than 1 implies negative association between itemsets. 

## Steps for association rules

Reading the data, converting the continuous into discrete data.

### Dataset operations

Single-dimension:
- Checking the distribution of most popular items
- Counting how many items were sold in general

Two-dimensions;
- What are the most popular pairs
- Measures and tests for pairs

### Single and two dimensions frequency inspection

Overview of multi-items sets
- selection according to support
- selection of most frequent

### Itemsets

Checking multi-items sets
- selection according to support 
- selection of most frequent items

### Rules

Here the patterns found in the dataset are converted into rules with causes and effects.

### Statistics

Jaccard dissimilarity makes cluster analysis available. 
