---
title: "International Economics dataset, association rules"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
author: "Mateusz Baryła"
date: "6 12 2020"
bibliography: bibliography.bibtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      results = 'asis')
pacman::p_load(tidyverse, kableExtra, kohonen, 
               arules, arulesCBA, arulesViz,
               randomNames)

knitr::opts_chunk$set(plain.ascii = FALSE,        # Always use this option in Rmd documents
           style        = "rmarkdown", # Always use this option in Rmd documents
           footnote     = NA,          # Makes html-rendered results more concise
           subtitle.emphasis = FALSE)  # Improves layout with some rmardown themes
```

## Introduction

Association rules are often used for reducing number of transactions in the 
databases where each row is coded by binary attributes. There is no need for 
them to be trained, or labeling dataset beforehand. Most common applications 
are found in the market basket analysis, discovering interesting patterns of 
DNA and protein sequences, common patterns of behavior can be found for 
customers that proceed customers dropping their cell phone operator.

## Dataset

### Instacart Market Basket Analysis

Firstly all the data need to be loaded to the R environment.

```{r}
aisles <- read.csv("data\\aisles.csv", sep=",")
departments <- read.csv("data\\departments.csv", sep=",")
order_products_train <- read.csv("data\\order_products__train.csv", sep=",")
orders <- read.csv("data\\orders.csv", sep=",")
products <- read.csv("data\\products.csv", sep=",")
random_names <- read.csv("E:\\GitHub\\unsupervised-learning\\data\\random_names.csv", sep=",")
```

Then in order to reduce the number of rows sample fraction out of order products 
is taken. We take 3% of transactions which still result in around 800 000 
transactions.

```{r}
set.seed(0402)
trans <- 
  order_products_train %>% 
  sample_frac(size = 0.5) %>% 
  inner_join(products, by = "product_id", keep = FALSE) %>%
  inner_join(aisles, by = "aisle_id", keep = FALSE) %>% 
  inner_join(orders, by = "order_id", keep = FALSE) %>% 
  inner_join(departments, by ="department_id", keep = FALSE)
```

Transactional data set is not very easy in interpretation in current format. 
As for now data set contains a lot of ids which are not very interpretable.
Since then a decision is made to perform several transformation which will 
result in transactional data.

```{r}
day_of_week <- 
  tibble(day_id = 0:6,
         dow_name = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
         )
reordered_df <- 
  tibble(reordered_id = 0:1,
         reorder_name = c("Not reordered", "Reordered")
         )
order_hour_of_day <- 
  tibble(
       hour_id = 0:23,
       time_of_day = c(rep("Night",6), rep("Morning", 4), rep("Senior hours", 2),
                    rep("Noon", 2), rep("Afternoon", 4), rep("Early night",4), rep("Night", 2))
       )
```

```{r}
trans_for_ar <- 
  trans %>% 
  inner_join(day_of_week, by = c("order_dow" = "day_id"), keep=FALSE) %>%
  inner_join(reordered_df, by = c("reordered" = "reordered_id"), keep=FALSE) %>%
  inner_join(order_hour_of_day, by = c("order_hour_of_day" = "hour_id"), keep=FALSE) %>%
  inner_join(random_names, by = c("user_id"= "user_id")) %>% 
  select(-aisle_id, -department_id, -eval_set,
         -product_id, -order_dow, - reordered, 
         -order_hour_of_day, -days_since_prior_order, 
         -order_number, -add_to_cart_order, -user_id) %>% 
  arrange(order_id, user_name)
```

Since the chosen data set is very large and only a sample of it has been taken 
into account the column days_since_prior_order does not make a lot of sense since
as for now not every user has order number 1 associated. Therefore there is 
a decision to omit this column as well order_number. 

```{r}
trans <- as(trans_for_ar, "transactions")
```

The final transactional dataframe is made of the following attributes:
* order_id - needed for an unique transaction specification
* product_name - full product name
* aisle - aisle name that identifies, where the product belongs inside the store
* department - there are 21 department names in the store
* dow_name - identifies day of week when order has been made
* reorder_name - whether the product has been reordered or not
* time_of_day - artificially created time of the day based on column value specification
* user_name - random user name assigned for easier identification than original user_id

# Association rules

## General view

During transactional data analysis the technique of association rules 
and mining frequent itemsets plays an  important role for retail basket analysis. 
This technique is especially useful in mining patterns inside large databases. 
What is the most frequently used are such statistics as frequent itemsets, 
maximal frequent itemsets, closed frequent itemsets and association rules. 

Definition:
Let I = {i1, i2, ..., in} be a set of n binary attributes called items. 
Let D = {t1, t2, ..., tn} be a set of transactions called database. Each 
transaction in D has a unique transaction ID and contains a subset of the items
in I. The sets of items X and Y are called antecedent and consequent of the rule.

The support(X) is defined as the count of transactions that contains item X to 
the number of transactions in the whole data set. 

The confidence of a rule is defined as conf(X => Y) = supp(X u Y)/supp(X). 

Another practical measure used for adressing issue of finding too many constraints
is measure called lift, whihc is defined as lift(X => Y)=supp(X u Y)/(supp(X)supp(Y)).
The most common interpretation is following, it can be interpreted as the deviation
of the support of the whole rule from the support expected under independence 
given the supports of the LHS and the RHS. 

## Associations mining

In order to investigate the nature of the dataset the read.transactions 
function has been used. 

```{r}
summary(trans)
```

```{r}
length(trans)
```

```{r}
size(trans)
```

## 

```{r}
# creating the rules - standard settings
rules.trans<-apriori(trans, parameter=list(supp=0.1, conf=0.5))

# sorting rules by confidence + displaying
rules.by.conf<-sort(rules.trans, by="confidence", decreasing=TRUE)
inspect(head(rules.by.conf))

rules.by.lift<-sort(rules.trans, by="lift", decreasing=TRUE) # sorting by lift
inspect(head(rules.by.lift))

rules.by.count<- sort(rules.trans, by="count", decreasing=TRUE) # sorting by count
inspect(head(rules.by.count))

rules.by.supp<-sort(rules.trans, by="support", decreasing=TRUE) 
inspect(head(rules.by.supp))
```

### Questions

## What is the profile of transactions that are being made during Senior hours?

```{r}
rules.Senior.hours<-apriori(data=trans, 
                               parameter=list(supp=0.001, conf=0.08), 
                               appearance=list(default="lhs", rhs= "Senior hours"), control=list(verbose=F))

rules.Senior.hours.byconf<-sort(rules.Senior.hours, by="confidence", decreasing=TRUE)
inspect(head(rules.Senior.hours.byconf))
```


Ones may want to figure out what an empty bracker on the lhs means. As documentation 
claims this rule means that no matter what other items are involved the item 
in the rhs will be chosen with the level of confidence which is equal to support.


## What is the profile of transactions that are being made during Senior hours?

```{r}
rules.Senior.hours<-apriori(data=trans, 
                               parameter=list(supp=0.001, conf=0.08), 
                               appearance=list(default="lhs", rhs= "Senior hours"), control=list(verbose=F))

rules.Senior.hours.byconf<-sort(rules.Senior.hours, by="confidence", decreasing=TRUE)
inspect(head(rules.Senior.hours.byconf))
```

We can clearly see that there is no difference for senior hours with ordering by
confidence and by lift. Given these metrics we spot that mostly beverages are chosen,
people have tendency to use Senior hours on Tuesdays. Others need additional coffeine.
It is clear that there are no aisle selected such as "eye ear care", "digestion"
which might suggest that these hours are often used by people with some serious disorders.
Let's try now to check what is the profile of these two categories based on hours.

```{r}
rules.muscles.joints.pain.relief<-apriori(data=trans, 
                               parameter=list(supp=0.0001, conf=0.000001), 
                               appearance=list(default="lhs", rhs= "muscles joints pain relief"), 
                                control=list(verbose=F))

rules.muscles.joints.pain.relief.bycoverage<-sort(rules.muscles.joints.pain.relief, by="coverage", decreasing=TRUE)
inspect(head(rules.muscles.joints.pain.relief.bycoverage))
```

As we can see senior hours are place in the sixts position but still coverage is very low.
For other metrics label that indicates these particular hoours has not been
placed in the top of the resulting table.

# How the Friday mood affects shoppings?

```{r}
rules.friday<-apriori(data=trans, 
                               parameter=list(supp=0.0001, conf=0.000001), 
                               appearance=list(default="lhs", rhs= "Friday"), 
                                control=list(verbose=F))

rules.friday.byconfidence<-sort(rules.friday, by="confidence", decreasing=TRUE)
inspect(head(rules.friday.byconfidence))
```

Alcohol is definitely placed very high, any sort of of it occurs in all of the 
top results. Once they add alcohol to the cart they do not reorder it. 
What might be interesting is that people usually buy alcohol 
in the morning. It might be simply because most of the customers on Friday 
do shopping in the morning in order to have more free time during the proper
weekend. 

```{r}
trans_for_ar %>% 
  group_by(dow_name, time_of_day) %>% 
  count() %>% 
  ungroup() %>% 
  group_by(dow_name) %>% 
  mutate(perc = n/sum(n))
```

The results shows that most of the purchases are being made on Friday afternoon.
That is interesting why people buy alcohol in the morning rather than in the afternoon.
What might also be absorbing is that people do not choose any salty food in
addition to the alcoholic drinks. 

```{r}
rules.lunch.meat<-apriori(data=trans, 
                               parameter=list(supp=0.0001, conf=0.000001), 
                               appearance=list(default="lhs", rhs= "lunch meat"), 
                                control=list(verbose=F))

rules.lunch.meat.byconfidence<-sort(rules.lunch.meat, by="confidence", decreasing=TRUE)
inspect(head(rules.lunch.meat.byconfidence))
```

```{r}
rules.lunch.meat<-apriori(data=trans, 
                               parameter=list(supp=0.0001, conf=0.000001), 
                               appearance=list(default="lhs", rhs= "lunch meat"), 
                                control=list(verbose=F))

rules.lunch.meat.byconfidence<-sort(rules.lunch.meat, by="confidence", decreasing=TRUE)
inspect(head(rules.lunch.meat.byconfidence))
```

Let's investigate particular weekdays in terms of some association  metrics.
```{r}
rules <- apriori(trans)
rules.sub <- subset(rules, subset = rhs %pin% "dow_name=")
```
