---
title: "Rules that inside shopping cart. How to instantly measure transactions relationships?"
author: "Mateusz Baryła"
date: "06.02.2021"
output:
  html_document:
    df_print: paged
    pandoc_args:
    - +RTS
    - -K512m
    - -RTS
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
bibliography: bibliography_association.bibtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      results = 'asis')
pacman::p_load(tidyverse, kableExtra, kohonen, 
               arules, arulesCBA, arulesViz,
               randomNames)

knitr::opts_chunk$set(plain.ascii = FALSE,
           style        = "rmarkdown",
           footnote     = NA,         
           subtitle.emphasis = FALSE,
           results = 'asis')
```

# Introduction

During last year a retail optimization has gained a lot of attention in 
the data science community. The COVID-19 pandemic resulted in restrictions
that brought restrictions that changed a behavior of customers. Retail 
industry forced the closure of physical stores. "The Next Normal" requires 
adaptation to new requirements and the winners are these retailers that 
priorities online retail. For instance in United States, e-commerce availability
and hygiene caused 17 percent of consumers to leave their primary store. One
of the American retailers, Instacart has already adressed issue of optimizing
e-commerce and released prediction competition on kaggle with several data sets [@kaggle]. 
**Instacart** is an online American retailer that uses advanced analytics algorithms
to provide best experience for their customers. 

# Dataset preprocessing

## Instacart Market Basket loading

Firstly all the data need to be loaded to the R environment. 

```{r}
aisles <- read.csv("data\\aisles.csv", sep=",")
departments <- read.csv("data\\departments.csv", sep=",")
order_products_train <- read.csv("data\\order_products__train.csv", sep=",")
orders <- read.csv("data\\orders.csv", sep=",")
products <- read.csv("data\\products.csv", sep=",")
random_names <- read.csv("E:\\GitHub\\unsupervised-learning\\data\\random_names.csv", sep=",")
```

Then in order to reduce the number of rows sample fraction out of order products
is taken. We take 50% of transactions which still result in more than half a million 
transactions.

```{r}
set.seed(0402)
trans <-
  order_products_train %>%
  sample_frac(size = 0.5) %>%
  inner_join(products, by = "product_id", keep = FALSE) %>%
  inner_join(aisles, by = "aisle_id", keep = FALSE) %>%
  inner_join(orders, by = "order_id", keep = FALSE) %>%
  inner_join(departments, by ="department_id", keep = FALSE)
```

## Preprocessing 

A transactional data set is not very easy in interpretation in current format.
As for now data set contains a lot of ids which are not very interpretable.
A decision has been made to perform several transformation which will
result in transactional data.

```{r}
kable_head <- function(df){
  df %>% kable() %>% head()
}
```

```{r}
day_of_week <-
  tibble(day_id = 0:6,
         dow_name = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
         )

day_of_week %>% kable_head

reordered_df <-
  tibble(reordered_id = 0:1,
         reorder_name = c("Not reordered", "Reordered")
         )

reordered_df %>% kable_head

order_hour_of_day <-
  tibble(
       hour_id = 0:23,
       time_of_day = c(rep("Night",6), rep("Morning", 4), rep("Senior hours", 2),
                    rep("Noon", 2), rep("Afternoon", 4), rep("Early night",4), rep("Night", 2))
       )

order_hour_of_day %>% kable_head
```

Above three mapping tables have been created. First two ones are self explanatory,
for hour of the day six time periods have been assigned for further analysis.

```{r}
trans_for_ar <-
  trans %>%
  inner_join(day_of_week, by = c("order_dow" = "day_id")) %>%
  inner_join(reordered_df, by = c("reordered" = "reordered_id")) %>%
  inner_join(order_hour_of_day, by = c("order_hour_of_day" = "hour_id")) %>%
  inner_join(random_names, by = c("user_id"= "user_id")) %>%
  select(-aisle_id, -department_id, -eval_set,
         -product_id, -order_dow, - reordered,
         -order_hour_of_day, -days_since_prior_order,
         -order_number, -add_to_cart_order, -user_id) %>%
  arrange(order_id, user_name)
```

Transactions are combined in one big dataframe, id columns are removed. Columns
that are also redundant in terms of analysis is eval_set. Since the chosen 
data set is very large and only a sample of it has been taken into account the 
column days_since_prior_order does not make a lot of sense since as for now not 
every user has order number 1 associated. Therefore there is a decision to omit 
this column as well order_number.

```{r, echo=FALSE}
rm(aisles, departments, order_products_train, orders, products, random_names,
   trans, day_of_week, order_hour_of_day, random_names)
```


```{r}
trans <- as(trans_for_ar, "transactions")
trans_for_viz <- as(trans_for_ar %>% select(product_name, aisle, department), "transactions")
```

The final transactional dataframe is made of the following attributes:

* order_id - needed for an unique transaction specification
* product_name - full product name
* aisle - aisle name that identifies, where the product belongs inside the store
* department - there are 21 department names in the store
* dow_name - identifies day of week when order has been made
* reorder_name - whether the product has been reordered or not
* time_of_day - artificially created time of the day based on column value specification
* user_name - artificial user name assigned for easier identification than original user_id

# Association rules

## General view

Association rules are often used for reducing number of transactions in the 
databases where each row is coded by binary attributes. There is no need for 
them to be trained nor to be labeled beforehand. Most common 
applications are found in the market basket analysis, discovering interesting patterns of 
DNA and protein sequences, common patterns of behavior can be found for 
customers that proceed customers dropping their cell phone operator.

During transactional data analysis the technique of association rules
and mining frequent itemsets plays an  important role for retail basket analysis.
This technique is especially useful in mining patterns inside large databases.
What is the most frequently used are such statistics as frequent itemsets,
maximal frequent itemsets, closed frequent itemsets and association rules.

Definition:
Let I = {i1, i2, ..., in} be a set of n binary attributes called items.
Let D = {t1, t2, ..., tn} be a set of transactions called database. Each
transaction in D has a unique transaction ID and contains a subset of the items
in I. The sets of items X and Y are called antecedent and consequent of the rule .

The support(X) is defined as the count of transactions that contains item X to
the number of transactions in the whole data set.

The confidence of a rule is defined as conf(X => Y) = supp(X u Y)/supp(X).

Another practical measure used for adressing issue of finding too many constraints
is measure called lift, whihc is defined as lift(X => Y)=supp(X u Y)/(supp(X)supp(Y)).
The most common interpretation is following, it can be interpreted as the deviation
of the support of the whole rule from the support expected under independence
given the supports of the LHS and the RHS [@JSSv014i15].

## Associations mining

In order to investigate the nature of the dataset the read.transactions
function has been used. Transformations lead to creating transactional matrix
with 565097 rows and 132889 columns. Most frequent items are not surprisingly
binary feature and transaction identificator which is order_id.
There are no missing values inside the database.

## Rules found in the dataset

```{r}
kableRules <- function(rules, sortable){
  rules.sorted<-sort(rules, by=sortable, decreasing=TRUE)
  rules.sorted.df <- as(head(rules.sorted), "data.frame")
  rownames(rules.sorted.df) <- NULL
  rules.sorted.df %>% 
    as_tibble() %>% 
    mutate(across(.cols = c(support, confidence, coverage, lift, count), 
                  ~round(., 4))) %>%
    kable()
}
```

```{r}
rules.trans<-apriori(trans, parameter=list(supp=0.1, conf=0.5),
                     control=list(verbose=FALSE))
```

Mined rules are specified with two threshold, with minima support 0.1 
and minimal confidence 0.5.

```{r}
kableRules(rules, "confidence")
```

All the listed items have confidence above 10.

```{r}
kableRules(rules, "lift")
```

Regarding lift we can 

```{r}
kableRules(rules, "count")
```

```{r}
kableRules(rules, "support")
```

# Questions

## Question 1

**What is the profile of transactions that are being made during Senior hours?**

```{r}
rules.Senior.hours<-apriori(data=trans,
                            parameter=list(supp=0.001, conf=0.08),
                            appearance=list(default="lhs", rhs= "time_of_day=Senior hours"),
                            control=list(verbose=F))
kableRules(rules.Senior.hours, "confidence")
```

Ones may want to figure out what an empty bracker on the lhs means. As documentation
claims this rule means that no matter what other items are involved the item
in the rhs will be chosen with the level of confidence which is equal to support.

## Question 2

**What is the profile of transactions that are being made during Senior hours?**

```{r}
rules.Senior.hours<-apriori(data=trans,
                            parameter=list(supp=0.001, conf=0.08),
                            appearance=list(default="lhs", rhs= "time_of_day=Senior hours"),
                            control=list(verbose=F))
kableRules(rules.Senior.hours, "lift")
```

We can clearly see that there is no difference for senior hours with ordering by
confidence and by lift. Given these metrics we spot that mostly beverages are chosen,
people have tendency to use Senior hours on Tuesdays. Others need additional coffeine.
It is clear that there are no aisle selected such as "eye ear care", "digestion"
which might suggest that these hours are often used by people with some serious disorders.
Let's try now to check what is the profile of these two categories based on hours.

```{r}
rules.muscles.joints.pain.relief<-apriori(data=trans,
                                          parameter=list(supp=0.0001, conf=0.000001),
                                          appearance=list(default="lhs",
                                                          rhs= "aisle=muscles joints pain relief"),
                                          control=list(verbose=F))
kableRules(rules.muscles.joints.pain.relief, "coverage")
```

As we can see senior hours are place in the sixts position but still coverage is very low.
For other metrics label that indicates these particular hoours has not been
placed in the top of the resulting table.

## Question 3

**How the Friday mood affects shoppings?**

```{r}
rules.friday<-apriori(data=trans,
                      parameter=list(supp=0.0001, conf=0.000001),
                      appearance=list(default="lhs", rhs= "dow_name=Friday"),
                      control=list(verbose=F))
kableRules(rules.friday, "confidence")
```

Alcohol is definitely placed very high, any sort of of it occurs in all of the
top results. Once they add alcohol to the cart they do not reorder it.
What might be interesting is that people usually buy alcohol
in the morning. It might be simply because most of the customers on Friday
do shopping in the morning in order to have more free time during the proper
weekend.

```{r}
trans_for_ar %>%
  group_by(dow_name, time_of_day) %>%
  count() %>%
  ungroup() %>%
  group_by(dow_name) %>%
  mutate(perc = n/sum(n))
```

The results shows that most of the purchases are being made on Friday afternoon.
That is interesting why people buy alcohol in the morning rather than in the afternoon.
What might also be absorbing is that people do not choose any salty food in
addition to the alcoholic drinks.

```{r}
rules.lunch.meat<-apriori(data=trans,
                               parameter=list(supp=0.0001, conf=0.000001),
                               appearance=list(default="lhs", rhs= "aisle=lunch meat"),
                                control=list(verbose=F))
kableRules(rules.lunch.meat, "confidence")
```

## Question 4

**Investigate weekdays in terms of some association  metrics.**

```{r}
rules <- apriori(trans, parameter=list(supp=0.001, conf=0.001))
rules.dow.tod <- subset(rules, subset = lhs %pin% "dow_name=" & rhs %pin% "time_of_day=")
kableRules(rules.dow.tod, "confidence")
```

In terms of first row of rules sorted by confidence 5.8 % of transactions
containing during Night has been made on Thursday. Discussing lift it
also indicates that the strongest association is between Thursday and Night.

```{r}
rules <- apriori(trans, parameter=list(supp=0.001, conf=0.001))
rules.tod.dow <- subset(rules, subset = lhs %pin% "time_of_day=" & rhs %pin% "dow_name=")
kableRules(rules.tod.dow, "count")
```

Most of the transactions found in dataset is on Monday Afternoon.

# Visualization techniques

## Hierarchical clustering

```{r}
trans.sel <-trans_for_viz[,itemFrequency(trans_for_viz)>0.05] # selected transations
d.jac.i<-dissimilarity(trans.sel, which="items") # Jaccard as default
plot(hclust(d.jac.i, method="ward.D2"), main="Dendrogram for items")
```

The interpretation of the hclust graph is following if we keep three biggest
clusters:
* the first group of transactions is formed out of fresh vegetables and fresh fruits
which do belong to produce department,
* the second group contains transactions made between 12 and 18,
* last but not least we have third cluster that can be best described as
mostly snacks and beverages which are bought either at night or between
10 and 12.

```{r}
itemFrequencyPlot(trans_for_viz, topN=10, type="absolute", main="Item Frequency")
```

In terms of items frequency it is shown that most of the record has been added
in the afternoon. Customers also focus on keeping in shape since a lot of
transactions are made inside produce department, others in dairy eggs.

```{r}
rules_for_viz <- apriori(trans_for_viz, parameter = list(support = 0.001))
plot(rules_for_viz, method = "group")
```

```{r}
plot(head(rules_for_viz, 30, by = "lift"), method = "paracoord", reorder =TRUE)
```

Figure above shows rules with arrows where the width of arrows is linked to support
and the color confidence contains information about confidence of given rule.
The values of each dimension are connected with each other via line. The y-axis
is formed via nominal values and the x-axis presents the position in the rule.

# Conclusion

Association rules are very useful technique for mining patterns in large datasets
due to their algorithms. What is more it can save a lot of time in finding user
preferences and what drives their choices. Most of the useful functionalities
are implemented either in a form of graphs or association rules mining functionalities.
For latter arulesViz is widely used where a lot of brand new visualization techniques to
research association rules has been implemented. The main features of arules in terms
of association rules are efficient implementation with sparse matrices usage.

# Bibliography
